---
title: "term7-project"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
#Load in the data
loans <- read.csv('prosperLoanData.csv')
library(ggplot2)
library(dplyr)

```
## Exploring the data

Here we will explore the data and choose some variables to examine in further depth.
```{r}
colnames(loans)

table(loans$LoanStatus)
```
I feel that the most interesting investigation into this data would be attempting to predict whether or not a loan defaults. This is obviously a tough thing, so I will first begin by examining the characteristics of defaulted loans versus completed ones.
```{r}
defaults <- subset(loans, LoanStatus == 'Defaulted')
completed <- subset(loans, LoanStatus == 'Completed')
```
The first things I am going to look at are credit grade and employment status. I expect that loans that default will have a lower credit grade and/or a more precarious employment status.
```{r}
ggplot(aes(x = CreditGrade), data = defaults) +
  geom_bar()
```
